The role of priors, competence and relevance for the interpretation of disjunction \thanks{We thank \ldots}

### 2021-11-25 writing/paper/paper_draft.tex
minor edits and comments on first write-up draft

@@

\subsection{Predictions}
Prior research has shown that the robustness of implicatures from ``some'' to ``some, but not all''  predicted by the standrard scalar implicature (SI) account is influenced by three factors. First, high contextual relevance of the stronger alternative ``all'' to the listener is predicted to render more robust scalar implicatures. Second, high speaker competence about the truth of the stronger alternative is predicted to lead to more robust scalar implicature generation. Lastly, high prior probability of the stronger ``all'' being true is assumed to \textit{decrease} the scalar implicature rate.
If the currently standard scalar implicature account also holds for the interpretation of disjunctions, analogous predictions for the robustness of the implicature from ``A or B'' to ``A or B, but not both'' can be derived. In case of disjunction, the relevant stronger alternative is ``and''.

@@

On critical trials, participants were asked to rate four sentences, one per factor and one containing an upper-bounded (exclusive) paraphrase of the trigger. The sentences elicited judgements how likely it is that each factor was high, and how likely the critical scalar implicature was true. The context stories and sentences manipulating the factors of interest were designed in an analogous fashion for both triggers and varied within-subjects, so the following descriptions apply for both triggers.

If our factor hypotheses are true, on average, we expect higher likelihood ratings for the statements eliciting relevance, competence and prior judgements given items in the respective high conditions. Critically, we expect higher likelihood ratings for the scalar implicature, if the relevance and comptenece in the item were judged as high and the prior of the alternative as low.
%Participants were then asked to rate statements eliciting four judgements using a slider bar: the relevance of the “all”/“and”-alternative, the competence of the speaker about the alternative, the prior probability of ”all” / “and” being true, and, crucially, the likelihood of the exclusive reading of “or” or of the upper-bounded reading of “some”.

\subsection{Procedure and Materials}
This study is a two-by-two-by-two-by-two within-subjects rating task (relevance $\times$ competence $\times$ prior $\times$ trigger), conducted as a web-based experiment. The study proceeds as follows.

On each trial, participants read a background story, followed by a sentence presented in a blue box which is meant to elicit the likelihood rating. These sentences were of the form ``It is important to X to know whether Y'' (relevance rating), ``Z knows whether Y'', ``If A, then B'' / ``If B , then A'' (disjunction prior rating), ``If some Y, then all Y'' (some prior rating), ``From what Z said we may conclude that W'', where X is the listener, Z is the speaker and Y is the target event in the background story, and W is the event under the upper bounded / exclusive reading. A and B are the disjuncts in case of disjunction. For the inference strength elicitation, participants additionally read a critical utterance of the form ``Z says to X: Y'', containing the trigger (``some'' / ``or''), presented below the background story in a red box.

@@

 We designed 32 different stories per trigger type (``some''~vs.~``or''), resulting in a total of 64 stories. Each subject sees eight randomly sampled stories (one per prior $\times$ competence $\times$ relevance condition out of four possible stories) such that they see four ``or'' and four ``some'' stories in randomized order. The assignment of conditions to the triggers is randomized between-subjects.

At the beginning, participants were welcomed to the experiment, read instructions and first saw three labelled example trials showing the working of the slider. They read a simple story about a person shopping. They then saw a question about the story that was clearly false (expected rating: 0), clearly true (expected rating: 100) and implied to be rather true (expected rating: 50-100), given the background story. These expected responses were labelled and explained.  
The main part of the experiment consisted of eight critical stories, randomly shuffled with eight attention check stories. The attention checks consisted of one trial, visually matching critical trials. On these trials, the vignettes contained a statement which clearly wrote out what participants were supposed to answer (e.g., ``Please move the slider maximally left'') in an area of text that participants had to read to complete the trial. Participants who failed more than two of these attention checks were excluded from the analysis. 
%Each critical story block has the following structure: Participants read a background story, followed by a comprehension statement which is either clearly true, clearly false, or uncertain (sampled at random for each story) which they are asked to rate using the slider. Then, they read statements presented in a box colored blue associated with the factors of interest and rate it using a slider bar. The factors of interest are relevance, competence and prior, in randomized order within-participants. In the “or”-condition, the prior statement consists of two symmetrical conditional sentences (see example below). Then, participants answer another three comprehension questions. These are followed by the critical utterance containing “or” / “some” appearing in a box colored red. Below, the target statement gathering the likelihood of the exclusive disjunction / the upper-bounded reading of “some” (i.e., the likelihood of the implicature) appears. Again, they rate the target statement using the slider. The background story remains on the screen throughout the trials. 
After the study, participants could fill out a socio-demographic questionnaire.
The experiment can be viewed under \texttt{https://magpie-xor-some.netlify.app/}. The preregistration for the experiment can be found under \texttt{https://osf.io/v7zjp}.

\subsection{Participants}
We recruited 277 participants through the crowd-sourcing platform Prolific. Participants were restricted to those whose first language included English, who previously took part in at least five other Prolific studies, and whose approval rate was at least 0.9, according to prescreening criteria of the platform. Participants took 20 minutes on average to complete the study and were compensated \pounds2.48 for their participation. Following our preregistered exclusion criteria, we excluded 15 participants for not indicating their native language, 3 participants for completing the study in under 8 minutes, 26 participants for failing more than 20\% of the attention checks, and 27 for failing more than 20\% of the comprehension questions. Due to a coding error of the comprehension questions, 206 participants were left post applying exclusion criteria, although the preregistered target sample size post exclusions was 200 subjects. We conducted a power analysis in order to determine the minimal sample size required for a power of 0.8 of finding evidence for the conjunction of all factor hypotheses (relevance, competence and prior) predicted by the SI account, finding that at least 180 subjects were required. 
In the following analyses, data from the 206 participants is analysed.\footnote{All analyses were also conducted on data from 200 subject, excluding the last six submissions. No noteworthy quantitative or qualitative differences of the results were observed.}

@@

\begin{figure*}[t]
	\begin{center}
		\includegraphics[width=1\linewidth]{images/posterior-effects-main.png}
	\end{center}
	\vspace{-0.3cm}
	\caption{Main analysis results: Distributions of the posterior samples of each factor of interest for the trigger ``some'' and ``or''. The colors and the vertical dashed lines indicate the effect intervalls (from left to right: negative, no, positive effect).}
	\label{posteriors-main}
\end{figure*}

Consistent with predictions, for the trigger ``some'', we find a clear negative prior probabilty effect of the stronger alternative ``all'' being true, as inidcated by the probability of the negative effect of prior being $P = 0.999$ (Fig.\ref{posteriors-main}, \texttt{prior\_some} line, orange color). Similarly, we find a clear positive effect for the effect of speaker competence ($P =  1$, Fig.\ref{posteriors-main}, \texttt{competence\_some} line, green color). However, we do not find any clear effects of relevance (Fig.\ref{posteriors-main}, \texttt{relevance\_some} line, split colors). If anything, the data supports the hypothesis that the effect is not negative.  
We hypothesize that the absense of a clear-cut relevance effect might be due to the difficulty manipulating this dimension contextually. It might be that the vignettes used in the experiment failed to make the listener relevance salient enough for any effects to be borne out. Further experiments should improve this manipulation, as well as additionally consider speaker relevance.

For the trigger ``or'', in contrast, we only find a positive effect of competence ($P =  0.993$, Fig.\ref{posteriors-main}, \texttt{competence\_xor} line, green color).  We find no comprehensive effects for relevance or competence. While the reason for the former might be the same as for ``some'', we suggest that these results might also be influenced by collinearity of the effects. We turn to this conjecture in light of the exploratory analyses, described below. 
Overall, these results would provide evidence against the scalar implicature hypothesis because not all factors yielded an effect on the robustness of scalar implicatures generated from ``some'' and ``or''. Similarly, the results would provide evidence against the identity hypothesis, as en effect of competence was observed for ``some'', but not for ``or''.

\subsection{Exploratory analysis}
\begin{figure*}[t]
	\begin{center}
		\includegraphics[width=1\linewidth]{images/posterior-effects-categorical.png}
	\end{center}
	\vspace{-0.3cm}
	\caption{Exploratory analysis with categorical predictors: Distributions of the posterior samples of each factor of interest for the trigger ``some'' and ``or''. The colors and the vertical dashed lines indicate the effect intervalls (from left to right: negative, no, positive effect).}
	\label{posteriors-cat}
\end{figure*}
In addition to the main analysis, we conducted two exploratory analyses. First, we fit a model using categorical predictors only. That is, we fit a Bayesian linear mixed effects regression model, regressing the inference rating against our prior intuitive classifications of the vignettes on the dimensions relevance, competence and prior (high or low, respectively), the effect of trigger, and their interactions. The random effects structure and the priors were identical to the main model. The factors were sum-coded, -1 representing the low condition, 1 representing the high condition. 
This model decouples the predictors by dichotomizing their classification and, therefore, addresses the potential collinearty issue. Furthermore, it provides a comparison between participants' judgements and our intuitive prior classifications of the vignettes.

Given this model and the data, we found very similar results for both triggers: We observe a negative effect of prior and a positive effect of competence, in line with predictions of the scalar implicature hypothesis (Fig.\ref{posteriors-cat}). However, we again observe no effects of relevance. If anything, the relevance effect tends in the direction opposite to predictions of the SI account for the trigger ``or''. 
\begin{figure*}[t]
	\begin{center}
		\includegraphics[width=1\linewidth]{images/posterior-effects-means.png}
	\end{center}
	\vspace{-0.3cm}
	\caption{Exploratory analysis with predictor ratings averaged by-vignette: Distributions of the posterior samples of each factor of interest for the trigger ``some'' and ``or''. The colors and the vertical dashed lines indicate the effect intervalls (from left to right: negative, no, positive effect).}
	\label{posteriors-mean}
\end{figure*}

These results match the second exploratory analysis, whereby the ratings were used as predictors in the linear model, but the ratings were averaged for each vignette across participants. In this second  exploratory model, we again find similar effects for ``some'' and ``or'': a positive effect of comptenece, negative for prior, and an indefinite relevance effect (Fig.\ref{posteriors-mean}). By averaging over single participant ratings, this model indicates that the absent competence effect in the main model might be due to by-participant variation.

In order to further investigate the relevance effects and potential collinarity concerns, we computed exploratory pairwise correlations of all the predictors. Interestingly, we find a significant correlation between the prior and the relevance effects for the trigger ``or'' ($R^2 = -0.106$). We also found a significant correlation between competence and relevance ($R^2 = 0.127$) for ``or''. No correlations were found for ``some''. These intriguing results might indicate that these effects are closely related for the disjunction. Intuitively, a negative correlation between prior and relevance of the stronger alternative ``and'' makes sense: the more likely the conjunction is a priori, the more likely it is that the listener might know about it already, yielding the implicature irrelevant to the listener. Similarly, competence might be connected to relevance indirectly via the prior: the more likely the stronger alternative is a priori, the more likely it is that a speaker knows about it, while the irrelevance still persists for the listener. However, this entangled results need further research.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% customization
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\usepackage[dvipsnames]{xcolor}
\newcommand{\mf}[1]{\textcolor{BurntOrange}{[MF: #1]}}
\newcommand{\pt}[1]{\textcolor{Cerulean}{[PT: #1]}}
\newcommand{\bvt}[1]{\textcolor{ForestGreen}{[BvT: #1]}}

@@

\subsection{Predictions}
Prior research has shown that the robustness of implicatures from ``some'' to ``some, but not all''  predicted by the standard scalar implicature (SI) account is influenced by three factors. First, high contextual relevance of the stronger alternative ``all'' to the listener is predicted to render more robust scalar implicatures. Second, high speaker competence about the truth of the stronger alternative is predicted to lead to more robust scalar implicature generation. Lastly, high prior probability of the stronger ``all'' being true is assumed to \textit{decrease} the scalar implicature rate.
If the currently standard scalar implicature account also holds for the interpretation of disjunctions, analogous predictions for the robustness of the implicature from ``A or B'' to ``A or B, but not both'' can be derived. In case of disjunction, the relevant stronger alternative is ``and''.

@@

On critical trials, participants were asked to rate four sentences, one per factor and one containing an upper-bounded (exclusive) paraphrase of the trigger. The sentences elicited judgements of how likely it is that each factor was high, and how likely the critical scalar implicature was true. The context stories and sentences manipulating the factors of interest were designed in an analogous fashion for both triggers and varied within-subjects, so the following descriptions apply for both triggers.
\mf{I suggest to give one full example of a vignette (context, critical questions, possibly test questions) for each trigger here; similar to what is now in the appendix (but with nicer formatting)}
\mf{it might be useful to have a graphical representation of the order of ``screens'' within each trial; e.g. similar to the example in Figure 1 here: \url{https://magpie-manual.netlify.app/00_getting_started/02_basics/}}

If our factor hypotheses are true, on average, we expect higher likelihood ratings for the statements eliciting relevance, competence and prior judgements given items in the respective high conditions. Critically, we expect higher likelihood ratings for the scalar implicature, if the relevance and comptenece in the item were judged as high and the prior of the alternative as low.
%Participants were then asked to rate statements eliciting four judgements using a slider bar: the relevance of the “all”/“and”-alternative, the competence of the speaker about the alternative, the prior probability of ”all” / “and” being true, and, crucially, the likelihood of the exclusive reading of “or” or of the upper-bounded reading of “some”.
\mf{Last paragraph feels redundant to me here; best repeat hypotheses in results / analyze section but not necessarily here}

\subsection{Procedure and Materials}
This study is a $ 2 \times 2 \times 2 \times 2$ within-subjects rating task (relevance $\times$ competence $\times$ prior $\times$ trigger), conducted as a web-based experiment. The study proceeds as follows.

On each trial, participants read a background story, followed by a sentence presented in a blue box which is meant to elicit the likelihood rating.\mf{maybe include a screenshot?} These sentences were of the form ``It is important to X to know whether Y'' (relevance rating), ``Z knows whether Y'', ``If A, then B'' / ``If B , then A'' (disjunction prior rating), ``If some Y, then all Y'' (some prior rating), ``From what Z said we may conclude that W'', where X is the listener, Z is the speaker and Y is the target event in the background story, and W is the event under the upper bounded / exclusive reading. A and B are the disjuncts in case of disjunction. For the inference strength elicitation, participants additionally read a critical utterance of the form ``Z says to X: Y'', containing the trigger (``some'' / ``or''), presented below the background story in a red box.

@@

 We designed 32 different stories per trigger type (``some''~vs.~``or''), resulting in a total of 64 stories. Each subject saw eight randomly sampled stories (one per prior $\times$ competence $\times$ relevance condition out of four possible stories) such that they saw four ``or'' and four ``some'' stories in randomized order. The assignment of conditions to the triggers is randomized between-subjects.

At the beginning, participants were welcomed to the experiment, read instructions and first saw three labelled example trials showing the working of the slider. They read a simple story about a person shopping. They then saw a question about the story that was clearly false (expected rating: 0), clearly true (expected rating: 100) and implied to be rather true (expected rating: 50-100), given the background story. These expected responses were labelled and explained.  
The main part of the experiment consisted of eight critical stories, randomly shuffled with eight attention check stories. The attention checks consisted of one trial, visually matching critical trials. On these trials, the vignettes contained a statement which clearly wrote out what participants were supposed to answer (e.g., ``Please move the slider maximally left'') in an area of text that participants had to read to complete the trial. Participants who failed more than two of these attention checks were excluded from the analysis. 
%Each critical story block has the following structure: Participants read a background story, followed by a comprehension statement which is either clearly true, clearly false, or uncertain (sampled at random for each story) which they are asked to rate using the slider. Then, they read statements presented in a box colored blue associated with the factors of interest and rate it using a slider bar. The factors of interest are relevance, competence and prior, in randomized order within-participants. In the “or”-condition, the prior statement consists of two symmetrical conditional sentences (see example below). Then, participants answer another three comprehension questions. These are followed by the critical utterance containing “or” / “some” appearing in a box colored red. Below, the target statement gathering the likelihood of the exclusive disjunction / the upper-bounded reading of “some” (i.e., the likelihood of the implicature) appears. Again, they rate the target statement using the slider. The background story remains on the screen throughout the trials. 
After the study, participants could voluntarily fill out a socio-demographic questionnaire.
The experiment can be viewed under \texttt{https://magpie-xor-some.netlify.app/}. The preregistration for the experiment can be found under \texttt{https://osf.io/v7zjp}.

\subsection{Participants}
We recruited 277 participants through the crowd-sourcing platform Prolific. Participants were restricted to those whose first language included English, who previously took part in at least five other Prolific studies, and whose approval rate was at least 0.9, according to prescreening criteria of the platform. Participants took 20 minutes on average to complete the study and were compensated \pounds2.48 for their participation. Following our preregistered exclusion criteria, we excluded 15 participants for not indicating their native language, 3 participants for completing the study in under 8 minutes, 26 participants for failing more than 20\% of the attention checks, and 27 for failing more than 20\% of the comprehension questions. Due to a coding error of the comprehension questions, 206 participants were left post applying exclusion criteria, although the preregistered target sample size post exclusions was 200 subjects. We conducted a power analysis in order to determine the minimal sample size required for a power of 0.8 of finding evidence for the conjunction of all factor hypotheses (relevance, competence and prior) predicted by the SI account, finding that at least 180 subjects were required.\mf{we should give some details about the power analysis or link to a document in the repository}
In the following analyses, data from the 206 participants is analysed.\footnote{All analyses were also conducted on data from 200 subject, excluding the last six submissions. No noteworthy quantitative or qualitative differences of the results were observed.}

@@

\begin{figure*}[t]
	\begin{center}
		\includegraphics[width=1\linewidth]{images/posterior-effects-main.png}
	\end{center}
	\vspace{-0.3cm}
	\caption{Main analysis results: Distributions of the posterior samples of each factor of interest for the trigger ``some'' and ``or''. The colors and the vertical dashed lines indicate the effect intervalls (from left to right: negative, no, positive effect). \mf{better use nicer labels for $y$-axis: use ``or'' not ``xor'' and no hyphens?; same for next plots}}
	\label{posteriors-main}
\end{figure*}

Consistent with predictions, for the trigger ``some'', we find a clear negative prior probabilty effect of the stronger alternative ``all'' being true, as inidcated by the probability of the negative effect of prior being $P = 0.999$ (Fig.\ref{posteriors-main}, \texttt{prior\_some} line, orange color). Similarly, we find a clear positive effect for speaker competence ($P =  1$, Fig.\ref{posteriors-main}, \texttt{competence\_some} line, green color). However, we do not find any clear effects of relevance (Fig.\ref{posteriors-main}, \texttt{relevance\_some} line, split colors). If anything, the data supports the hypothesis that the effect is not negative.
We hypothesize that the absense of a clear-cut relevance effect might be due to the difficulty manipulating this dimension contextually. It might be that the vignettes used in the experiment failed to make the listener relevance salient enough for any effects to be borne out. Further experiments should improve this manipulation, as well as additionally consider speaker relevance.
\mf{best first just state the results and use the last two sentences for ``Discussion'' section}

For the trigger ``or'', in contrast, we only find a positive effect of competence ($P =  0.993$, Fig.\ref{posteriors-main}, \texttt{competence\_xor} line, green color).  We find no comprehensive effects for relevance or competence. While the reason for the former might be the same as for ``some'', we suggest that these results might also be influenced by collinearity of the effects. We turn to this conjecture in light of the exploratory analyses, described below. 
Overall, these results would provide evidence against the scalar implicature hypothesis because not all factors yielded an effect on the robustness of scalar implicatures generated from ``some'' and ``or''. Similarly, the results would provide evidence against the identity hypothesis, as an effect of competence \mf{prior?} was observed for ``some'', but not for ``or''.\mf{we should explain somewhere where the threshold of 0.05 comes from; when discussing the results here we could add that (think) no effects becomes credible even if the threshold would be reduced to 0}

\subsection{Exploratory analysis}
\begin{figure*}[t]
	\begin{center}
		\includegraphics[width=1\linewidth]{images/posterior-effects-categorical.png}
	\end{center}
	\vspace{-0.3cm}
	\caption{Exploratory analysis with categorical predictors: Distributions of the posterior samples of each factor of interest for the trigger ``some'' and ``or''. The colors and the vertical dashed lines indicate the effect intervalls (from left to right: negative, no, positive effect). \mf{I'm not sure if we should blindly reuse a threshold of 0.05 here without further justification. Something to think about.}}
	\label{posteriors-cat}
\end{figure*}
In addition to the main analysis, we conducted two exploratory analyses. First, we fit a model using categorical predictors only. That is, we fit a Bayesian linear mixed effects regression model, regressing the inference rating against our prior intuitive classifications of the vignettes on the dimensions relevance, competence and prior (high or low, respectively), the effect of trigger, and their interactions. The random effects structure and the priors were identical to the main model. The factors were sum-coded, -1 representing the low condition, 1 representing the high condition. 
This model decouples the predictors by dichotomizing their classification and, therefore, addresses the potential collinearty issue. Furthermore, it provides a comparison between participants' judgements and our intuitive prior classifications of the vignettes.

Given this model and the data, we found very similar results for both triggers: We observe a negative effect of prior and a positive effect of competence, in line with predictions of the scalar implicature hypothesis (Fig.\ref{posteriors-cat}). However, we again observe no effects of relevance. If anything, the relevance effect tends in the direction opposite to predictions of the SI account for the trigger ``or''. 
\begin{figure*}[t]
	\begin{center}
		\includegraphics[width=1\linewidth]{images/posterior-effects-means.png}
	\end{center}
	\vspace{-0.3cm}
	\caption{Exploratory analysis with predictor ratings averaged by-vignette: Distributions of the posterior samples of each factor of interest for the trigger ``some'' and ``or''. The colors and the vertical dashed lines indicate the effect intervalls (from left to right: negative, no, positive effect). \mf{I'm not sure if we should blindly reuse a threshold of 0.05 here without further justification. Something to think about.}}
	\label{posteriors-mean}
\end{figure*}

These results match the second exploratory analysis, whereby the ratings were used as predictors in the linear model, but the ratings were averaged for each vignette across participants. In this second  exploratory model, we again find similar effects for ``some'' and ``or'': a positive effect of competence, negative for prior, and an indefinite relevance effect (Fig.\ref{posteriors-mean}). By averaging over single participant ratings, this model indicates that the absent competence effect in the main model might be due to by-participant variation.

In order to further investigate the relevance effects and potential collinarity concerns, we computed exploratory pairwise correlations of all the predictors. Interestingly, we find a significant correlation between the prior and the relevance effects for the trigger ``or'' ($R^2 = -0.106$). We also found a significant correlation between competence and relevance ($R^2 = 0.127$) for ``or''. No correlations were found for ``some''. These intriguing results might indicate that these effects are closely related for disjunction. Intuitively, a negative correlation between prior and relevance of the stronger alternative ``and'' makes sense: the more likely the conjunction is a priori, the more likely it is that the listener might know about it already, yielding the implicature irrelevant to the listener. Similarly, competence might be connected to relevance indirectly via the prior: the more likely the stronger alternative is a priori, the more likely it is that a speaker knows about it, while the irrelevance still persists for the listener. However, this entangled results need further research.
